VarSight: Prioritizing Clinically Reported Variants with Binary Classification Algorithms

### 2019-01-17 paper/main.tex
updating .gitignore and adding Julies edits

\abstract{\textbf{Motivation:} In genomic medicine for rare disease patients, the major goal is to identify one or more variants that cause disease.  Typically, this is done through filtering and then prioritization of variants for manual curation.  However, prioritization of variants in rare disease patients remains a challenging task due to the high degree of variability in phenotype presentation and molecular source of disease.  Thus, methods that can identify and/or prioritize variants to be clinically reported in the presence of such variability of critical importance.  \\
\textbf{Results:} We tested the application of classification algorithms that ingest variant predictions along with phenotype information for predicting whether a variant will be ultimately clinically reported and returned to a patient.  To test the classifiers, we performed a retrospective study on variants that were clinically reported to patients in the Undiagnosed Diseases Network.  We measured the classifier against another variant prioritization algorithm along with two single-measure controls to show how the classifiers tend to outperform the other methods.  The tested classifiers had a median rank of 7.0-8.0 for all variants and 2.0-3.5 for variants reported as pathogenic.  \\
\textbf{Availability:} The scripts used to generate results are available at \href{https://github.com/HudsonAlpha/VarSight}{https://github.com/HudsonAlpha/VarSight}. \\
\textbf{Contact:} \href{jholt@hudsonalpha.org}{jholt@hudsonalpha.org}\\
\textbf{Supplementary information:} Supplementary data are available at \textit{Bioinformatics}
online.}

@@

\subsection{Alignment and variant calling}
After sequencing, we followed GATK best practices \citep{depristo2011} to align to the hg19 reference with BWA-mem \citep{li2013}.  Aligned sequences we processed via GATK for base quality score recalibration, indel realignment, and duplicate removal.  Finally, SNV and indel variants were joint genotyped, again according to GATK best practices \citep{depristo2011}.  The end result of this pipeline is a Variant Call Format (VCF) file used in the following sections.

@@

Codicem also performs filtering for the analysts to reduce the number of variants that are viewed through a standard clinical analysis.  We used the latest version of their primary clinical filter to replicate the standard filtering process of the analysts.  In short, the filter requires the following for a variant to pass through the clinical filter: sufficient total read depth, sufficient alternate read depth, low population frequency, at least one predicted effect on a transcript, at least one gene-disease association, and to not be a known, common false-positive from sequencing.  In general, the filter reduces the number of variant from the order of millions to hundreds (anecdotally, roughly 200-400 variants per proband after filtering).  For the specific details on the filter used, please refer to the metadata available in the VarSight repository.

@@

The first method uses base annotations provided by the HPO to calculate a simple cosine score \citep{kohler2017} between the patient's phenotypes and each gene.  This method tends to be more conservative because it relies solely on curated annotations from the HPO.  The second method, a tool called PyxisMap \citep{wilk2018}, uses the same annotations from the HPO, but adds in automatically text-mined data from NCBI's PubTator \citep{wei2013} and performs a Random-Walk with Restart \citep{page1999} on the ontology graph structure.  The PyxisMap method has the added benefit of incorporating fresh gene-phenotype connections from newer papers, but it also tends to make more spurious connections due to the imprecision of the text-mining from PubTator.  Each method generates a single numerical feature that is used in the following analyses.

\subsection{Patient selection}
In our analysis, we focused on variants that were clinically reported as ``primary", meaning the analysts believed the variant to be directly related to the patient's phenotype.  Note that secondary and/or incidental findings are specifically not included in this list.  The analysts assigned each primary variant a rank from variant of uncertain significance (VUS), likely pathogenic, or pathogenic based on the recommendations in the ACMG guidelines for variant classification \citep{richards2015}.

@@

\subsection{Model training and tuning}
As noted earlier, there are generally hundreds of variants per proband that make it through the filter, but only a few are ever clinically reported.  As a result, there is a major imbalance in the number of true positives (variants clinically reported) and true negatives (variants seen, but not clinically reported).  Across all 237 proband samples, there were a total of 378 clinically reported variants and another 87819 variants that were seen but not reported.  We split the data into training and test sets on a per-proband basis with the primary goal of roughly balancing the total number of true positives in each set.  Additionally, the cases were assigned to a particular set by chronological order of analysis in order to reduce any chronological biases that may be introduced by expanding scientific knowledge (i.e. there are roughly equal proportions of ``early" or ``late" proband samples from the UDN in each set).  In the training set, there were a total of 189 returned variants and 44593 not returned variants spanning 120 different probands.  In the test set, there were a total of 189 returned variants and 43226 not returned variants spanning 117 different probands.  In our results, the returned test variants are further stratified in their reported levels of pathogenicity.

@@

For comparison, we selected to run Exomiser \citep{smedley2015a} because it performed comparatively well with all phenotypes even in the presence of noise (see benchmarking from \citealp{smedley2015b}).  We used followed the installation on their website to install Exomiser CLI v.11.0.0 along with version 1811 for hg19 data sources.  For each test case, we created a VCF file from the pre-filtered list of annotated variants created by Codicem and passed those VCFs along with the full list of patient HPO terms to Exomiser.  We then parsed the output JSON file created by Exomiser into a ranked order of variants.  Of note, Exomiser did not rank every variant that was present in the filtered VCF, typically ranking only 20-50 variants from the pre-filtered VCF.  For primary variants that were unranked by Exomiser, we conservatively assumed that the Exomiser ranked those variants as just outside the range of what was returned (e.g. if Exomiser ranked 15 variants, we assumed a missing variant was ranked 16th).

@@

\section{Conclusion}
We assessed the application of binary classification algorithms for identifying variants that were ultimately reported on a clinical report.  We trained and tested these algorithms using real patient variants and phenotype terms from the standard clinical process used for the Undiagnosed Diseases Network (UDN).  From a classification perspective, we found that these methods tend to have low precision scores, meaning a high number of false positives were identified by each method.  However, when evaluated as a ranking system, all four methods out-performed single-measure ranking systems and Exomiser.  We consider the best classifier to be the EasyEnsembleClassifier from {\it imblearn} that had a median rank of 7.0 for all reported variants while ranking 73\% in the top 20 for the case.  For ``Pathogenic" variants, the median rank was 2.0 and 97\% of those variants were ranked in the top 20 for the case.  While not perfect, the ability for the algorithms to push reported variants to very early in the list is quite promising.

We expect these classification algorithms could be improved or refined in a variety of ways.  First, adding new features and/or removing unused features could lead to improvements in the algorithm.  The top two most important features were both phenotype related suggesting that a highly accurate phenotype scoring system would greatly benefit these algorithms.  Many features had a low importance, so pruning them may improve the overall results.  Additionally, some of the features represent data that is not freely available to the research community, so pruning or replacing those features with publicly accessible sources would likely influence the results.  Second, there may be a better classification algorithm for this type of data.  The four selected classifiers were all freely available methods intended to handle the large class imbalance in the training set, but other algorithms that aren't as readily available may improve the result.

Ultimately, we believe the trained classifiers in VarSight are a significant step forward in handling real-world data for rare, undiagnosed disease patients.  Despite the variability in phenotype and allowance of variants of uncertain significance in reporting, the models trained in VarSight still ranked reported variants highly, especially variants labeled as ``Pathogenic".  Going forward, we expect the standard for variant ranking systems will behave similar to those in VarSight, fully integrating all available phenotype and variant annotations into one model.

\section*{Acknowledgements}
We are grateful for the participation of patients and family members and all of their referring clinicians. We would like to acknowledge all of the teams within the UDN including the coordinating center and all of the clinical sites working hard to provide definitive diagnoses for UDN patients. We would also like to acknowledge the UDN WGS core headed by Dr. Shawn Levy.

@@

%HPO paper to cite
\bibitem[Köhler {\it et~al}., 2018]]{kohler2018} Köhler, Sebastian, \textit{et~al.} "Expansion of the Human Phenotype Ontology (HPO) knowledge base and resources." {\it Nucleic acids research} (2018).

@@

%PubTator
\bibitem[Wei {\it et~al}., 2013]]{wei2013} Wei, Chih-Hsuan, Hung-Yu Kao, and Zhiyong Lu. "PubTator: a web-based text mining tool for assisting biocuration." {\it Nucleic acids research} 41.W1 (2013): W518-W522.

>>>>

\abstract{\textbf{Motivation:} In genomic medicine for rare disease patients, the primary goal is to identify one or more variants that cause disease.  Typically, this is done through filtering and then prioritization of variants for manual curation.  However, prioritization of variants in rare disease patients remains a challenging task due to the high degree of variability in phenotype presentation and molecular source of disease.  Thus, methods that can identify and/or prioritize variants to be clinically reported in the presence of such variability are of critical importance.  \\
\textbf{Results:} We tested the application of classification algorithms that ingest variant predictions along with phenotype information for predicting whether a variant will ultimately be clinically reported and returned to a patient.  To test the classifiers, we performed a retrospective study on variants that were clinically reported to patients in the Undiagnosed Diseases Network.  We measured the classifier against another variant prioritization algorithm along with two single-measure controls to show how the classifiers tend to outperform the other methods.  The tested classifiers had a median rank of 7.0-8.0 for all variants and 2.0-3.5 for variants reported as pathogenic.  \\
\textbf{Availability:} The scripts used to generate results are available at \href{https://github.com/HudsonAlpha/VarSight}{https://github.com/HudsonAlpha/VarSight}. \\
\textbf{Contact:} \href{jholt@hudsonalpha.org}{jholt@hudsonalpha.org}\\
\textbf{Supplementary information:} Supplementary data are available at \textit{Bioinformatics}
online.}

@@

\subsection{Alignment and variant calling}
After sequencing, we followed GATK best practices \citep{depristo2011} to align to the hg19 reference with BWA-mem \citep{li2013}.  Aligned sequences were processed via GATK for base quality score recalibration, indel realignment, and duplicate removal.  Finally, SNV and indel variants were joint genotyped, again according to GATK best practices \citep{depristo2011}.  The end result of this pipeline is a Variant Call Format (VCF) file used in the following sections.

@@

Codicem also performs filtering for the analysts to reduce the number of variants that are viewed through a standard clinical analysis.  We used the latest version of the primary clinical filter to replicate the standard filtering process of the analysts.  In short, the filter requires the following for a variant to pass through the clinical filter: sufficient total read depth, sufficient alternate read depth, low population frequency, at least one predicted effect on a transcript, at least one gene-disease association, and to not be a known, common false-positive from sequencing.  In general, the filter reduces the number of variant from the order of millions to hundreds (anecdotally, roughly 200-400 variants per proband after filtering).  For the specific details on the filter used, please refer to the metadata available in the VarSight repository.

@@

The first method uses base annotations provided by the HPO to calculate a simple cosine score \citep{kohler2017} between the patient's phenotypes and each gene.  This method tends to be more conservative because it relies solely on curated annotations from the HPO.  The second method, a tool called PyxisMap \citep{wilk2018}, uses the same annotations from the HPO, but adds in automatically text-mined data from NCBI's PubTator \citep{wei2013} and performs a Random-Walk with Restart \citep{page1999} on the ontology graph structure.  The PyxisMap method has the added benefit of incorporating gene-phenotype connections from recent papers that have not been manually curated into the HPO, but it also tends to make more spurious connections due to the imprecision of the text-mining from PubTator.  Each method generates a single numerical feature that is used in the following analyses.

\subsection{Patient selection}
In our analysis, we focused on variants that were clinically reported as ``primary", meaning the analysts believed the variant to be directly related to the patient's phenotype.  Note that secondary and/or incidental findings are specifically not included in this list.  The analysts assigned each primary variant a classification from variant of uncertain significance (VUS), likely pathogenic, or pathogenic based on the recommendations in the ACMG guidelines for variant classification \citep{richards2015}.

@@

\subsection{Model training and tuning}
As noted earlier, there are generally hundreds of variants per proband that pass the filter, but only a few are ever clinically reported.  Across all 237 proband samples, there were a total of 378 clinically reported variants and another 87819 variants that were seen but not reported.  As a result, there is a major imbalance in the number of true positives (variants clinically reported) and true negatives (variants seen, but not clinically reported).

We split the data into training and test sets on a per-proband basis with the primary goal of roughly balancing the total number of true positives in each set.  Additionally, the cases were assigned to a particular set by chronological order of analysis in order to reduce any chronological biases that may be introduced by expanding scientific knowledge (i.e. there are roughly equal proportions of ``early" or ``late" proband samples from the UDN in each set).  In the training set, there were a total of 189 returned variants and 44593 not returned variants spanning 120 different probands.  In the test set, there were a total of 189 returned variants and 43226 not returned variants spanning 117 different probands.  In our results, the returned test variants are further stratified in their reported levels of pathogenicity.

@@

For comparison, we selected to run Exomiser \citep{smedley2015a} because it performed comparatively well with all phenotypes even in the presence of noise (see benchmarking from \citealp{smedley2015b}).  We followed the installation on their website to install Exomiser CLI v.11.0.0 along with version 1811 for hg19 data sources.  For each test case, we created a VCF file from the pre-filtered list of annotated variants created by Codicem and passed those VCFs along with the full list of patient HPO terms to Exomiser.  We then parsed the output JSON file created by Exomiser into a ranked order of variants.  Of note, Exomiser did not rank every variant that was present in the filtered VCF, typically ranking only 20-50 variants from the pre-filtered VCF.  For primary variants that were unranked by Exomiser, we conservatively assumed that the Exomiser ranked those variants as just outside the range of what was returned (e.g. if Exomiser ranked 15 variants, we assumed a missing variant was ranked 16th).

@@

\section{Conclusion}
We assessed the application of binary classification algorithms for identifying variants that were ultimately reported on a clinical report.  We trained and tested these algorithms using real patient variants and phenotype terms from the standard clinical process used for the Undiagnosed Diseases Network (UDN).  From a classification perspective, we found that these methods tend to have low precision scores, meaning a high number of false positives were identified by each method.  However, when evaluated as a ranking system, all four methods out-performed single-measure ranking systems and Exomiser.  We consider the best classifier to be the EasyEnsembleClassifier from {\it imblearn} that had a median rank of 7.0 for all reported variants while ranking 73\% in the top 20 for the case.  For ``Pathogenic" variants, the median rank was 2.0 and 97\% of those variants were ranked in the top 20 for the case.  While these algorithms are not perfect classifiers, their use as a prioritization system is quite promising.

We expect these classification algorithms could be refined in a variety of ways.  First, adding new features and/or removing unused features could lead to improvements in the algorithm.  The top two most important features were both phenotype related suggesting that a highly accurate phenotype scoring system would greatly benefit these algorithms.  Many features had a low importance, so pruning them may improve the overall results.  Additionally, some of the features represent data that is not freely available to the research community, so pruning or replacing those features with publicly accessible sources would likely influence the results.  Second, there may be a better classification algorithm for this type of data.  The four selected classifiers were all freely available methods intended to handle the large class imbalance in the training set, but other algorithms that aren't as readily available may improve the result.

We believe the trained classifiers in VarSight are a significant step forward in reducing the problem of variant fatigue.  The models improve our ability to prioritize variants despite the variability and uncertainty injected by real-world data.  Ultimately, we believe implementing these models will enable analysts to assess the best candidate variants first, reducing the time to analyze a case and return molecular diagnoses to patients.

\section*{Acknowledgements}
We are grateful for the participation of patients and family members and all of their referring clinicians. We would like to acknowledge all of the teams within the UDN including the coordinating center and all of the clinical sites working hard to provide definitive diagnoses for UDN patients. We would also like to acknowledge the UDN whole genome sequencing core headed by Dr. Shawn Levy.

@@

%HPO paper to cite
\bibitem[Köhler {\it et~al}., 2018]{kohler2018} Köhler, Sebastian, \textit{et~al.} "Expansion of the Human Phenotype Ontology (HPO) knowledge base and resources." {\it Nucleic acids research} (2018).

@@

%PubTator
\bibitem[Wei {\it et~al}., 2013]{wei2013} Wei, Chih-Hsuan, Hung-Yu Kao, and Zhiyong Lu. "PubTator: a web-based text mining tool for assisting biocuration." {\it Nucleic acids research} 41.W1 (2013): W518-W522.
